Toronto Shelter Analysis: A Deep Dive of  Shelter Total Decedents with Features in Real Life

### 2024-09-27 paper/paper.qmd
update the sketchs, llm usage, dicussion part, and paper pdf

---
title: "My title"
subtitle: "My subtitle if needed"
author: 
  - Colin Sihan Yang
thanks: "Code and data are available at: https://github.com/hoodiexxx/deaths_shelter."
date: today
date-format: long
abstract: "This study explores the seasonal patterns and the deaths of the homeless people who are under the roof of the shelter system in Toronto. The research encompasses the deaths among the shelter residents and conducts analysis for 2007-2024 basing on open data material received from Open Data Toronto. It does have a particular diligence when winter months regime with death increase is noted while also establishing the effect of the COVID-19 pandemic on the ever rising death rates between the years 2019 and 2024. The results underscore the need for targeted health strategies particularly related to medical, and psychiatric services and shelter improvements in vulnerable periods."
format: pdf
toc: true
number-sections: true
bibliography: references.bib
---

@@

The increasing number of homeless people deaths in the City of Toronto has emerged as a core public health concern which portrays the harsh situation of the unaccommodated people. According to the studies that were recently published by the Toronto Public Health [@citeTPH],79 homeless persons were reported to die in the first six months of 2023, many of the deaths due to diease, flu, and epidemic that has plagued the city [@citeTorontoMeida]. This mortality not only implicates the health consequences of diease, but it also indicates systemic deficits in provision of safe facilities and healthcare services.

The increasing rates of homelessness coupled with the complexity of health issues faced by this population present a critical challenge for urban centers worldwide. Toronto, Canada's largest city, exemplifies this challenge, with a significant homeless population reliant on the shelter system. Recent trends and emerging data suggest that the mortality rate among these individuals is influenced by both environmental factors, such as harsh winter conditions, and systemic pressures including public health crises like the COVID-19 pandemic, drawing on open data from the period 2007 to 2024 provided by Open Data Toronto.[@citeTPH]. It also demonstrates that there is an apparent gap when it comes to devising public health policies to manage the risks posed by drug misuse in a matter of urgency and address the key requirements of housing and Epidemiological sanitary conditions in the long term[@citeTPH].

A mix of plans like the Toronto Overdose Action Plan or the Integrated Prevention and Harm Reduction Initiative (iPHARE) aimed at minimization of these risks have been adopted by the city. Collectively, they have provided some harm reduction services and addressed some mental health needs of their clients as well. Although, the high mortality rates have remained persistent and there is a need to contextualize these strategies and more efforts from the provincial and federal governments to protect both the measures used in the city and those who need the city most.

@@

The data used in this paper is sourced from Open Data Toronto and is load into this paper through the opendatatoronto library [@opendatatorotno]. The particular data set is used to analyze the deaths of Toronto shelter residents[@citeshelterdeaths]. All the data analysis was done through R [@citeR] with the aid of the following packages: tidyverse [@citetidyverse], here [@citehere], tinytex [@citetinytex], dplyr [@citedplyr], tibble [@citetibble], janitor[@citejanitor], ggplot2 [@citeggplot2], and knitr [@citeknitr].

The data used is published by Toronto Shelter & Support Services, providing information about the deaths of shelter residents. The data used in this paper is up-to-date as of Sept 19, 2024, the raw data set features the total decedents with 212 observations and 7 attributes. The dataset includes the id, Year, Month, Total Decedents, Male, Female, Transgender/Non-Binary/two_spirit. A sample of the cleaned data can be seen in @tbl-cleaned-data and a bar grapgh showcasing every observation, by date of collection, can be seen in @fig-lead-samples-over-time.

@@

the clean data has 212 rows x 3 columns(year, month, total_decedents)

```{r}
#| label: tbl-cleaned-data
#| tbl-cap: Sample of cleaned lead data 
#| echo: false

@@

cleaned_data |>
  ggplot(aes(x = "", y = total_decedents)) +
  geom_boxplot() +
  labs(y = "total_decedents", x = "") + 
  theme_classic()
```

@@

## First discussion point {#sec-first-point}

If my paper were 10 pages, then should be be at least 2.5 pages. The discussion is a chance to show off what you know and what you learnt from all this.

## Second discussion point

## Third discussion point

## Weaknesses and next steps

Weaknesses and next steps should also be included.

@@

# Additional data details

>>>>

---
title: "Toronto Shelter Analysis: A Deep Dive of  Shelter Total Decedents with Features in Real Life"
author: 
  - Colin Sihan Yang
thanks: "Code and data are available at: https://github.com/hoodiexxx/deaths_shelter."
date: today
date-format: long
abstract: "This study explores the seasonal patterns and the deaths of the homeless people who are under the roof of the shelter system in Toronto. The research encompasses the deaths among the shelter residents and conducts analysis for 2007-2024 basing on open data material received from Open Data Toronto. It does have a particular diligence when winter months regime with death increase is noted while also establishing the effect of the COVID-19 pandemic on the ever rising death rates between the years 2019 and 2024. The results underscore the need for targeted health strategies particularly related to medical, and psychiatric services and shelter improvements in vulnerable periods."
format: pdf
toc: true
number-sections: true
bibliography: references.bib
---

@@

The increasing number of homeless people deaths in the City of Toronto has emerged as a core public health concern which portrays the harsh situation of the unaccommodated people. According to the studies that were recently published by the Toronto Public Health [@citeTPH], 79 homeless persons were reported to die in the first six months of 2023, many of the deaths due to diease, flu, and epidemic that has plagued the city [@citeTorontoMeida]. This mortality not only implicates the health consequences of diease, but it also indicates systemic deficits in provision of safe facilities and healthcare services.

The increasing rates of homelessness coupled with the complexity of health issues faced by this population present a critical challenge for urban centers worldwide. Toronto, Canada's largest city, exemplifies this challenge, with a significant homeless population reliant on the shelter system. Recent trends and emerging data suggest that the mortality rate among these individuals is influenced by both environmental factors, such as harsh winter conditions, and systemic pressures including public health crises like the COVID-19 pandemic, drawing on open data from the period 2007 to 2024 provided by Open Data Toronto[@citeTPH]. It also demonstrates that there is an apparent gap when it comes to devising public health policies to manage the risks in a matter of urgency and address the key requirements of housing and Epidemiological sanitary conditions in the long term [@citeTPH].

@@

The data used in this paper is sourced from Open Data Toronto and is load into this paper through the opendatatoronto library [@opendatatorotno]. The particular data set is used to analyze the deaths of Toronto shelter residents [@citeshelterdeaths]. All the data analysis was done through R [@citeR] with the aid of the following packages: tidyverse [@citetidyverse], here [@citehere], tinytex [@citetinytex], dplyr [@citedplyr], tibble [@citetibble], janitor [@citejanitor], ggplot2 [@citeggplot2], and knitr [@citeknitr].

The data used is published by Toronto Shelter & Support Services, providing information about the deaths of shelter residents. The data used in this paper is up-to-date as of Sept 19, 2024, the raw data set features the total decedents with 212 observations and 7 attributes. The dataset includes the id, Year, Month, Total Decedents, Male, Female, Transgender/Non-Binary/two_spirit. A sample of the cleaned data can be seen in @tbl-cleaned-data and a bar grapgh showcasing every observation, by date of collection, can be seen in @fig-clean-data-year-month-decedents.

@@

the clean data has 212 rows x 3 columns (year, month, total_decedents)

```{r}
#| label: tbl-cleaned-data
#| tbl-cap: Sample of cleaned shelter death data 
#| echo: false

@@

cleaned_data |>
  ggplot(aes(x = "", y = total_decedents)) +
  geom_boxplot() +
  labs(y = "total decedents", x = "") + 
  theme_classic()
```

@@

## Major Conclusions and Their Effects

The evaluation of Toronto shelter decedent data in the years 2007–2024 makes it possible to expose the significance of environmental and public health factors in the overall mortality rates among the homeless population. One critical point is the recurring death rates when there is a persistent and severe cold, especially in January and February. There is a clear justification about the level of health and shelter provision that is needed during these months and it provides health services and shelter that have better insulations and heaters among other therapeutic care.

The COVID-19 pandemic has changed mostly attributed to the suffering increases which caused deaths in the shelter decedent data during 2019-2024 which is evident. This correlates with wider public health data with respect to covid where unfortunate incidence bashed on the covid vulnerable people. A set of conditions comprising homelessness, lower health standards and elevated incidence of Covid 19 created a health emergency such as is demonstrated by this dataset.

## Health Services and Seasonal Strategies

The seasonal variation of mortality rates in Toronto shelters complements the rationale for public health intervention. As per the results, policy options should be directed at increasing healthcare systems support at the end of the winter season. This might include deploying more mobile health units, expanding the emergency shelter system, and assuring that seasonal health related services such as flu shots and respiratory health clinical services are in ready at the location of the shelter throughout the cold months.

Moreover, shelter conditions should be improved against hostile environments especially winter season. Insulating the shelters and providing sufficient heat is important as if not many fatalities would be from avoidable reasons such as cold exposure. Achieving the trends discussed and recommended in this analysis would give the bridge that includes the ways of ensuring that the government makes appropriate expenditure during the vulnerable months and saves lives in the process.

## Influence of the COVID–19 Pandemic

It was clear from the analysis, that there was a surge in deaths within the shelter facilities, especially, during the years 2019 – 2024, which was unprecedented within the analysis period. The increase in shelter deaths during these periods as at the apex of the pandemic indicates that, these health services could not provide assistance to these people in the best way. In the middle of a pandemic, the homeless were placed in a precarious situation, living in congregate settings, requiring and often not getting healthcare, and having multiple chronic conditions – this was the conditions that COVID19 catalyzed. The data demonstrates the lessons that should be learned regarding the management of public health interventions at the onset of protest movements in order to leverage better leadership at the beginning of the protests.

From here-on, it is imperative that the long term plans for the provision of public health and shelter services in Toronto includes pandemic preparedness. This would be about guaranteeing that shelter establishments are built to allow for social distancing, that stockpiling of PPE, sanitation supplies and equipment is done, and that there is a well-organized infrastructure to conduct medical intervention for the homeless people such as vaccination campaigns at the invocation of a public health emergency.

## Weaknesses and Next Steps

it has to be said that this study is not without limitations. Most importantly, the dataset does not provide specifics on the individual healthy condition, causes of mortality, and other interfacing attributes such as surrounding conditions (e.g. temperatures and so on, along the same lines as shelter conditions).” In the analysis presented here, future development would make it possible by doing detailed death field causation or institutional studies of correlating temperature data to deaths on raising the fatality during extreme cold days.

Further, the dataset does not review what measures were taken by the city in the way of the interventions under review. Posterity may consider the tracking of some public health interventions and the betterment of shelters as variables accounting the variations in deaths to assess the impact of various strategies aimed at the cutting down of deaths. In addition, comparisons of data on shelter decedents from other cities where stronger measures such as providing additional cold weather shelters have been implemented may help to clarify which interventions are effective in achieving which results.

Finally, among the factors related to homelessness and its consequences – disease and premature death of the homeless – the comprehensive research would include the investigation of substance and mental health services optimization. Exploring how the realization of access to psychiatric and addiction treatment is beneficial on the mortality rates would be an additional tool on how to minimize the threats of shelter residents.

In summary, this study makes it evident that it is absolutely necessary to implement target, seasonal and health measures in order to lower mortality rates in shelters located within the city of Toronto. There are indeed lessons to be learnt from the impact of the COVID-19 pandemic in terms of preparedness and adaptability that can be applied in the future public health strategies targeting the vulnerable population.

@@
